---
title: "CAPE results"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: show
    collapsed: no
    toc: yes
    toc_float: yes
params:
    results_dir: here("tests", "testthat", "results")
---

Set the working directory below and run this workflow to see figures of results
with explanations.

```{r results_dir}
library(here);library(cape);library(knitr);library(DT)
results_dir <- params$results_dir
```

## Introduction

Here are the results of your CAPE analysis. This package implements a method, 
originally described in [Carter et al. (2012)](https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1003010) and [Tyler et al. (2013)](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1003270), that infers directed interaction networks 
between genetic variants for predicting the influence of genetic perturbations 
on quantitative traits. This method takes advantage of complementary information 
in partially pleiotropic genetic variants to resolve directional influences 
between variants that interact epistatically.

## Trait Decomposition

The first step in CAPE is often to perform a singular value decomposition (SVD) 
on the trait matrix. This step decomposes the traits into decorrelated, composite 
traits called eigentraits. The reasoning behind this step is that correlated 
components of related traits will be concentrated into individual eigentraits
thereby making these components easier to map. 

We plot this decomposition to show how each trait contributes to each eigentrait,
as well as the variance explained by each eigentrait. 

In the following plot, traits are shown along the $y$-axis and eigentraits
are shown along the $x$-axis. Each cell of the heat map indicates how much 
each trait contributes to each eigentrait. Positive contributions are shown 
in brown and negative contributions are shown in blue. Darker colors indicate 
stronger contributions. This heat map allows us to see which traits are compared 
or contrasted in the individual eigentraits. 

The gray bars along the top of the heat map indicate how much overall trait 
variance is explained by each eigentrait. They are always listed in order of 
the amount of variance explained. 

```{r svd, results = "asis", echo = FALSE}
fig.file <- file.path(results_dir, "svd.jpg")
fig.line <- paste0("![Eigentraits](", fig.file, "){width=70%}\n")
cat(fig.line)
```

## Single Locus Effects {.tabset .tabset-fade .tabset-pills}

Before performing the pairwise marker scans, we assess the association between
each individual marker and each trait or eigentrait using marker regression. 
These results are shown in the plots below.

The singlescan plots show how strongly each marker is associated with each 
trait. The strength of these main effects can be uses as a marker-selection
criterion for performing the pair-wise marker scans. When 
*marker_selection_method* is set to *top_effects*, the markers with the 
largest main effects are selected for the pair-wise marker scan regardless of 
significance. 

The single-locus effects are displayed in two different types of plots. 
One plot shows the overall locus effect, as well as the individual allele 
effects. For a two-parent cross, these are identical. However, for a multi-parent
cross, such as the Diversity Outbred (DO) or Collaborative Cross (CC) mice,
the allele effects will show how each parental haplotype affects the trait. 

In CAPE we use one parental haplotype as the reference. Thus all haplotype 
effects shown are relative to that reference. The reference in a DO/CC population
is typically the B6 (B) allele. 

The effects in the plots below are shown as lines, but this, and many other 
plotting parameters can be changed in *plot_singlescan()*.


```{r singlescan_effects, results = "asis", echo = FALSE}
fig.files <- list.files(path = results_dir, pattern = "Effects.jpg")
fig.names <- gsub(".jpg", "", fig.files)
for(i in 1:length(fig.files)){
    cat("###", fig.names[i], "\n")
    fig.line <- paste0("![", fig.names[i], "](", file.path(results_dir, fig.files[i]), ")\n\n")
    cat(fig.line)
}
```


## Standardized Single Locus Effects {.tabset .tabset-fade .tabset-pills}


The other way we display the single-locus effects is as standardized effects, 
which shows how significant each effect is, rather than the size and the 
direction of the effects. These standardized effects are shown below.

```{r singlescan_std, results = "asis", echo = FALSE}
fig.files <- list.files(path = results_dir, pattern = "Standardized.jpg")
fig.names <- gsub(".jpg", "", fig.files)
for(i in 1:length(fig.files)){
    cat("###", fig.names[i], "\n")
    fig.line <- paste0("![", fig.names[i], "](", file.path(results_dir, fig.files[i]), ")\n\n")
    cat(fig.line)
}
```


## Variant Influences 

The Combined Analysis of Pleiotropy and Epistasis results in a network
of directed genetic interactions. These results are plotted in multiple
different ways. 

### Variant influences as a matrix

One way to show the directed influences is in a matrix. In the following
plot, genetic markers are listed both on the $x$-axis and the $y$-axis.
The $x$-axis shows the markers as sources. Genetic interactions are 
directed from these source markers to the target markers on the $y$-axis. 
They are shown as either positive (brown) and negative (blue). Main effects
are shown along the right-hand side of the plot in extra columns.
These main effects line up with the source markers. The main effects
are also shown as positive (brown), or negative (blue). Main effects
are shown in terms of the original traits if *transform_to_phenospace*
is set to TRUE, or the eigentraits if *transform_to_phenospace* is set
to FALSE.

By default, cape does not test marker pairs that are correlated with a
Pearson $R >= 0.5$. Entries for pairs that are not tested are shown in 
gray. Gray blocks indicate areas of highly correlated markers. This 
matrix type of plot is good for seeing overall patterns of interactions
for dense matrices, or small numbers of markers. It is not very good for
visualizing sparse matrices derived from large numbers of markers.

```{r variant_influences, results = "asis", echo = FALSE}
fig.file <- file.path(results_dir, "variant_influences.jpg")
fig.line <- paste0("![variant influences](", fig.file, "){width=110%}\n")
cat(fig.line)
```


### Circular network plot

Genetic interactions can also be visualized in a circular plot. In these plots,
the chromosomes are plotted in a circle. The sizes are not representative of 
the true chromosome sizes. Main effects are shown in concentric circles around
the chromosomes, one trait per circle. Main effects are depicted as colored 
points or bars in this plot. For example, a brown bar at chromosome 7 for 
trait 1 means that there is a locus on chromosome 7 with a positive main 
effect on trait 1.

Genetic interactions in this plot are shown as colored arrows between 
genetic locations. Arrows show the direction of the influence, and the 
color indicates whether the source variant enhances (brown) or suppresses
(blue) the effects of the target marker. 

Main effects and interaction effects are also shown for covariates, which
are plotted in the circle before chromosome 1. 

This type of plot is good for seeing patterns in the physical location of 
genetic interactions in sparse networks. It is not good for visualizing 
networks with many interactions.

```{r circular_network, results = "asis", echo = FALSE}
fig.file <- file.path(results_dir, "Network_Circular.jpg")
fig.line <- paste0("![Circular Network](", fig.file, ")\n")
cat(fig.line)
```


### Network view

Finally, the cape interaction network is shown in a more classic network view. 
In this view, each genetic marker is depicted as a circle. Its main effects
are shown in slices of the pie as either positive (brown) or negative (blue).

Interactions are shown as arrows between linkage blocks. We cluster markers
into linkage blocks based on pairwise correlation. Network nodes are labeled
by chromosome followed by a number indicating the linkage block. Linkage blocks
can contain one or more markers. Covariates are also included in these network 
views, and have bold outlines. 

This type of plot is good for seeing patterns in network connectivity, such as 
clusters of enhancing or suppressing interactions, or nodes that serve
as network hubs. Genomic position information is lost in this view and thus 
this type of network is not good for visualizing the physical location of 
interacting markers. 

```{r network_layout, results = "asis", echo = FALSE}
fig.file <- file.path(results_dir, "Network_View.jpg")
fig.line <- paste0("![Network Layour](", fig.file, ")\n")
cat(fig.line)
```

## Variant influences tables {.tabset .tabset-fade .tabset-pills}

Another useful output of cape is the variant influences table. This table
contains all the specific information about interactions up to an adjusted
$p$ value of 0.2. The run_cape() pipeline writes two tables: 
Variant_Influences.csv and Variant_Influences_Interactions.csv.

The first table, Variant_Influences.csv, contains both genetic interactions
and main effects. Interactions are listed by their source marker and target 
marker. If the effect listed is a main effect, the target listed is a trait
rather than a marker. Because all effects are assessed from pairwise models,
even the main effects are listed with a conditioning marker. The main effect
of a marker is the largest effect that marker had conditioned on another 
marker. The conditioning marker and its positional information are listed 
in the case of main effects. Covariates can also have interactions and 
main effects. In this case, they are listed as the target our source marker.

These tables can be used to identify and plot specific interactions.
They are sorted by standardized effect size. Interactions and main effects 
can be plotted using the function plot_effects().

The following tables show the variant influences both with and without
main effects. 

### All Influences

```{r var_inf_table, echo = FALSE}
var.inf.table <- as.matrix(read.csv(file.path(results_dir, "Variant_Influences.csv"), stringsAsFactors = FALSE))
var.inf.table[,1] <- gsub("X", "Chr", var.inf.table[,1])
var.inf.table[,4] <- gsub("X", "Chr", var.inf.table[,4])
datatable(var.inf.table)
```

### Just Interactions

```{r var_int_table, echo = FALSE}
var.int.table <- as.matrix(read.csv(file.path(results_dir, "Variant_Influences_Interactions.csv"), stringsAsFactors = FALSE))
var.int.table[,1] <- gsub("X", "Chr", var.int.table[,1])
var.int.table[,4] <- gsub("X", "Chr", var.int.table[,4])
datatable(var.int.table)
```


## Plotting individual interactions {.tabset .tabset-fade .tabset-pills}

The following plots show the trait effects of the top interaction in 
your cape analysis. Different plots show the same data using bars,
lines, and points. 

If you are using continuously valued "genotypes," for example if you 
are using transcript levels as exlanatory variables instead of genotypes,
you can use plot_type = "h" to show heatmaps of predicted trait values
as the explanatory variables vary continuously.


```{r read_effects}
data_obj <- readRDS(file.path(results_dir, "cross.RDS"))
geno_obj <- readRDS(file.path(results_dir, "cross_geno.RDS"))
var.inf <- as.matrix(read.csv(file.path(results_dir, "Variant_Influences_Interactions.csv"), stringsAsFactors = FALSE))
```

### Line Plots

Line plots are a standard way to show genetic interactions. They show the 
median and error for individuals in each genotype group. One marker is plotted
along the $x$-axis, and the other is plotted with different line styles.

```{r line_plot}
plot_effects(data_obj, geno_obj, var.inf[1,"Source"], var.inf[1,"Target"], 
plot_type = "l", error_bars = "se", pheno_type = "eig", gen_model = "Recessive")
```

### Bar Plots

Bar plots also show summary statistics for each group defined by 
pairwise genotype. Each bar shows the median trait value for the 
group. Error bars, if specified with the error_bars argument, can 
show the standard deviation ("sd") or the standard error ("se"). 
Numbers over the bars indicate how many individuals are in each group.

```{r bar_plot}
plot_effects(data_obj, geno_obj, var.inf[1,"Source"], var.inf[1,"Target"],
plot_type = "b", error_bars = "se", pheno_type = "eig", gen_model = "Recessive")
```

### Point Plots
Point plots show all data points divided up into groups based on the 
pairwise genotypes. Horizontal line segments show the median value for 
each group.

```{r point_plot}
plot_effects(data_obj, geno_obj, var.inf[1,"Source"], var.inf[1,"Target"],
plot_type = "p", error_bars = "se", pheno_type = "eig", gen_model = "Recessive")
```


## Modifying effect plots {.tabset .tabset-fade .tabset-pills}

For each of these plots, you can specify which markers you want to 
see the effects of. You can also specify whether you would like to 
see their effects on eigentraits, normalized phenotypes, or raw 
traits, by changing the argument *pheno_type*. Normalized traits 
are plotted by default.

The genotype coding ("gen_coding" can also be specified to show 
additive effects ("Additive"), dominant effects ("Dominant"),
or recessive effects ("Recessive"). 

### Additive Model

In the **additive model**, genotypes are coded as shown in the following table.
In this model, the effect of the alternate allele is related to dosage. Having two
copies has a greater effect than one copy.

The plots below might not look very good depending on how many traits you 
are plotting. We encourage you to modify the plotting code to plot other 
interactions, other trait types, and to change the plot dimensions.

```{r, add_table, echo = FALSE}
add_table <- data.frame(cbind(c(0, 0.5, 1), 
c("Homozygous Reference", "Heterozygous", "Homozogous Alternate"), c(0, 1, 2)))
colnames(add_table)  <- c("Genotype Value","Genotype", "# Copies of Alternate Allele")
kable(add_table)
```

### Dominant Model

In the **dominant model**, genotypes are coded as shown in the following
table. To have an effect the alternate allele can be present in either one
or two copies.

```{r, dom_table, echo = FALSE}
dom_table <- data.frame(cbind(c(0, 1), c("Homozygous Reference", "Heterozygous and Homozygous Alternate"), c(0, "1 or 2")))
colnames(dom_table)  <- c("Genotype_Value", "Genotype", "# Copies of Alternate Allele")
kable(dom_table)
```

### Recessive Model

In the **recessive model**, genotypes are coded as follows. To have an effect, the 
alternate allele must be present in two copies. 

```{r, rec_table, echo = FALSE}
rec_table <- data.frame(cbind(c(0, 1), c("Homozygous Reference and Heterozygous", "Homozygous Alternate"), c("0 or 1", 2)))
colnames(rec_table)  <- c("Genotype_Value", "Genotype", "# Copies of Alternate Allele")
kable(rec_table)
```

## Getting help

Here we describe each output generically. For help interpreting specific 
results, or generating specific plots, please email <anna.tyler@jax.org>.
